---
title: "Sensitivity Analysis"
subtitle: "Assigment #3 Profit with Almond Yield"
date: "2025-04-21"
output:
  html_document:
    theme: cerulean
author: Kaiju Morquecho
---


### Load libraries
```{r message=FALSE}
library(here)
library(tidyverse)
library(ggpubr)
```

### Read data
```{r}
clim <- read.table(here("data","clim.txt"),header = TRUE) # read in climate data 
```

### Prepare data
```{r}
# model uses climate data from a particular month, must transform data by summarizing daily data into monthly data


# TEMPERATURE
feb_tmin <- clim %>% 
  
  # filter to month of interest
  filter(month == 2) %>%
  
  # group by year
  group_by(year) %>% 
  
   # calculate mean temp in february by year, and set column name to match function parameter
  summarize(feb_tmin = mean(tmin_c, na.rm = TRUE),
    .groups = "drop") 


# PRECIPITATION
jan_precip <- clim %>%
  
  # filter to month of interest
  filter(month == 1) %>%  

  # group by year 
  group_by(year) %>%
  
  
   # calculate sum of precip in january by year, and set column name to match function parameter
  summarize(
    jan_precip = sum(precip, na.rm = TRUE),
    .groups = "drop")

# bring together to have one df to pass into function 

clim <- left_join(feb_tmin, jan_precip, by = "year")


 
```

### Source functions and test new profit and wrapper functions 
```{r}
source(here("almond_profit_model.R"))

source(here("almond_anomaly.R"))

source(here("almond_profit.R"))

clim <- almond_profit_model(clim, base_yield = 1.0, price = 4000)

glimpse(clim)
```

### Run sensitivity analysis varying two parameters
```{r}
# source functions (if haven't already)
source(here("almond_profit_model.R"))

source(here("almond_anomaly.R"))

source(here("almond_profit.R"))


# sample size 
reps_per_year <- 10
years <- clim$year
n <- length(years) * reps_per_year # to make plotting nicer 

# simulate parameter uncertainty for feb_tmin, jan_precip and base_yield
set.seed(40)

sensitivity_analysis_df <- tibble(
  feb_tmin = rnorm(n, mean = 7, sd = 1),
  jan_precip = runif(n, min = 30, max = 600),
  base_yield = rnorm(n, mean = 1.1, sd = 0.1)
)

# apply almond_profit_model using pmap
sensitivity_outputs <- sensitivity_analysis_df %>%
  pmap_dfr(function(feb_tmin, jan_precip, base_yield) {
    inputs <- tibble(feb_tmin=feb_tmin, jan_precip=jan_precip)
    almond_profit_model(inputs, base_yield = base_yield)
    
    
  })

```

### Create plots
```{r}
# sensitivity_outputs$year <- sample(clim$year, size = nrow(sensitivity_outputs), replace = TRUE)

# changed year col to have more reps so boxplot would be clear and easy to understand patterns  
sensitivity_outputs$year <- rep(clim$year, each = reps_per_year)


# yield anomaly by year 
ggplot(sensitivity_outputs, aes(x = as.factor(year), y = yield_anomaly, group = year)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Yield Anomaly by Year (with Parameter Uncertainty)",
    x = "Year",
    y = "Yield Anomaly (tons/acre)"
  ) +
  theme_minimal()

ggplot(sensitivity_outputs, aes(x = yield_anomaly, y = profit, color = jan_precip)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_viridis_c(option = "C", name = "Jan Precip (mm)") +
  labs(
    title = "Yield Anomaly vs Profit (Colored by January Precipitation)",
    x = "Yield Anomaly (tons/acre)",
    y = "Profit (USD/acre)"
  ) +
  theme_minimal()

```

```{r}
p1 <- ggplot(sensitivity_outputs, aes(jan_precip, yield_anomaly)) +
  geom_point(alpha = 0.6, color = "springgreen3") +
  labs(x = "January Precip (mm)", y = "Yield Anomaly")

p2 <- ggplot(sensitivity_outputs, aes(feb_tmin, yield_anomaly)) +
  geom_point(alpha = 0.6, color = "indianred2") +
  labs(x = "Feb Tmin (°C)", y = NULL)


p3 <- ggplot(sensitivity_outputs, aes(jan_precip, profit)) +
  geom_point(alpha = 0.6, color = "gold2") +
  scale_y_continuous(
    breaks = scales::pretty_breaks(n = 10),  # aim for ~10 breaks
    labels = scales::dollar_format(prefix = "$", accuracy = 1)
  ) +
  labs(
    title = "Profit vs January Precipitation",
    x = "January Precipitation (mm)",
    y = "Profit per Acre (USD)"
  ) +
  theme_minimal()


ggarrange(p1, p2 , ncol = 2)

p3
```


The almond yield anomaly and profit model highlights the effect of climate variables, primarily that of precipitation , on yield outcomes almonds. There is a strong nonlinear relationship between january precipitation and the yield anomalies seen, where yield rapidly increases as precipitation rises. Temperature shows a more scattered relationship with yield anomaly, its effect is less pronounced and perhaps more complex than the model can account for. The sensitivity analysis, in summary, shows the benefits of modeling parameter uncertainty, not just as a way of testing functions and models' effectiveness and fit, but also as a way to understand the system or process that we are modeling.